---
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
dir.create("bench/figures", recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(
  fig.path = "bench/figures/readme-",
  fig.width = 7,
  fig.height = 4
)
```

# R Multi-Threaded Interpreter Experiment (`mtlapply`)

This repo is an experimental R runtime that can evaluate *pure R* closures concurrently on multiple OS threads, in a single R process, using `mtlapply()` (modeled after `lapply()`/`mclapply()`).

The hook is simple: when your workload is "a lot of independent R work" (feature engineering, per-shard transforms, per-group summaries), `mtlapply()` can give near-linear speedups without forking and without serializing return values.

## A Benchmark You Can Read (And Reproduce)

The "unit of parallelism" here is a shard id. There is no up-front "pre-splitting"; the `mtlapply()` call is what shards the work.

```{r benchmark-code, eval=FALSE}
# This is the actual benchmark code (identical in shape to bench/readme_bench_run.R).
# The key idea: mtlapply() shards work by shard id; the worker computes its own
# row indices deterministically. No up-front "split" required.

N <- 2e6L
nshards <- 64L
ngroups <- 4096L
feat_loops <- 40L

# Deterministic data (no RNG, no strings).
x <- (as.double(seq_len(N) %% 1000L) - 500) / 10
y <- (as.double((seq_len(N) * 17L) %% 1000L) - 500) / 10
w <- (as.double((seq_len(N) * 31L) %% 1000L) + 1) / 1000
grp <- rep_len(seq_len(ngroups), N)

worker <- function(shard_id) {
  idx <- seq.int(shard_id, N, by = nshards)
  z <- x[idx]
  for (i in seq_len(feat_loops)) {
    z <- log1p(abs(z)) + sin(y[idx] + z) * w[idx] + cos(z - y[idx])
  }
  g <- grp[idx]  # groups for summarise
  # summarise: group-wise sum, plus counts
  n <- tabulate(g, ngroups)
  s <- numeric(ngroups)
  for (j in seq_along(z)) {
    gj <- g[[j]]
    s[[gj]] <- s[[gj]] + z[[j]]
  }
  list(sum = s, n = n)
}

reduce <- function(parts) {
  s <- Reduce(`+`, lapply(parts, `[[`, "sum"))
  n <- Reduce(`+`, lapply(parts, `[[`, "n"))
  s / n
}

ids <- seq_len(nshards)

# baseline
system.time(reduce(lapply(ids, worker)))[["elapsed"]]

# parallel (in the experimental build)
system.time(reduce(mtlapply(ids, worker, threads = 8L)))[["elapsed"]]
```

## Real Numbers (Loaded From Artifacts)

The benchmark is run as a standalone base-R script under:

- the system `R` (to check for single-threaded regressions)
- `./build-mtl/bin/R` from this tree (to measure `mtlapply()` scaling)

Generate the timing artifacts:

```sh
mkdir -p bench/results

# system R (no mtlapply)
R --vanilla -q -f bench/readme_bench_run.R --args bench/results/system.rds

# R-devel (no mtlapply)
/usr/local/bin/R-devel --vanilla -q -f bench/readme_bench_run.R --args bench/results/rdevel.rds

# experimental build (has mtlapply)
./build-mtl/bin/R --vanilla -q -f bench/readme_bench_run.R --args bench/results/mtl.rds

# optional: generate a bench::mark artifact under the experimental build
R_LIBS_USER=/private/tmp/mtl-proof-lib-MW1vv9 R_LIBS_SITE='' \
  ./build-mtl/bin/R --vanilla -q -f bench/readme_bench_mark.R --args bench/results/mtl_bench_mark.rds
```

Then render this README, which loads those artifacts and summarizes them:

```{r results, echo=FALSE}
sys_path <- "bench/results/system.rds"
rdevel_path <- "bench/results/rdevel.rds"
mtl_path <- "bench/results/mtl.rds"
bm_path <- "bench/results/mtl_bench_mark.rds"

missing <- c(sys_path, rdevel_path, mtl_path, bm_path)[
  !file.exists(c(sys_path, rdevel_path, mtl_path, bm_path))
]
if (length(missing)) {
  stop(
    "missing benchmark artifacts:\n- ",
    paste(missing, collapse = "\n- "),
    "\n\nRun bench/readme_bench_run.R (and bench/readme_bench_mark.R) first."
  )
}

sys <- readRDS(sys_path)
rdevel <- readRDS(rdevel_path)
mtl <- readRDS(mtl_path)
bm <- readRDS(bm_path)

stopifnot(isTRUE(all.equal(sys$meta$settings, rdevel$meta$settings)))
stopifnot(isTRUE(all.equal(sys$meta$settings, mtl$meta$settings)))
stopifnot(isTRUE(all.equal(sys$meta$workloads, rdevel$meta$workloads)))
stopifnot(isTRUE(all.equal(sys$meta$workloads, mtl$meta$workloads)))

tab <- rbind(
  transform(sys$results, build = "system"),
  transform(rdevel$results, build = "rdevel"),
  transform(mtl$results, build = "mtl")
)

tab$label <- ifelse(tab$method == "lapply",
                    "lapply",
                    paste0("mtlapply(", tab$threads, ")"))

baseline_sys <- sys$results[sys$results$method == "lapply", c("workload", "median_seconds")]
baseline_rdevel <- rdevel$results[rdevel$results$method == "lapply", c("workload", "median_seconds")]
baseline_mtl <- mtl$results[mtl$results$method == "lapply", c("workload", "median_seconds")]
names(baseline_sys)[[2L]] <- "baseline_sys_lapply"
names(baseline_rdevel)[[2L]] <- "baseline_rdevel_lapply"
names(baseline_mtl)[[2L]] <- "baseline_mtl_lapply"

tab <- merge(tab, baseline_sys, by = "workload", all.x = TRUE, sort = FALSE)
tab <- merge(tab, baseline_rdevel, by = "workload", all.x = TRUE, sort = FALSE)
tab <- merge(tab, baseline_mtl, by = "workload", all.x = TRUE, sort = FALSE)

tab$speedup_vs_sys_lapply <- tab$baseline_sys_lapply / tab$median_seconds
tab$speedup_vs_rdevel_lapply <- tab$baseline_rdevel_lapply / tab$median_seconds
tab$speedup_vs_mtl_lapply <- tab$baseline_mtl_lapply / tab$median_seconds
tab$efficiency_vs_mtl_lapply <- ifelse(tab$method == "mtlapply",
                                       tab$speedup_vs_mtl_lapply / tab$threads,
                                       NA_real_)

tab_plot <- tab

tab <- tab[order(tab$workload, tab$build, tab$method, tab$threads), c(
  "workload", "build", "label", "median_seconds",
  "speedup_vs_rdevel_lapply", "speedup_vs_sys_lapply",
  "speedup_vs_mtl_lapply", "efficiency_vs_mtl_lapply"
)]

settings <- sys$meta$settings
settings_tab <- data.frame(
  setting = names(settings),
  value = vapply(settings, function(x) paste(x, collapse = ","), character(1)),
  stringsAsFactors = FALSE
)

r_versions <- data.frame(
  build = c("system", "rdevel", "mtl"),
  r_version = c(sys$meta$r_version, rdevel$meta$r_version, mtl$meta$r_version),
  stringsAsFactors = FALSE
)

cat("Settings:\n")
knitr::kable(settings_tab, row.names = FALSE)

cat("\nR versions:\n")
knitr::kable(r_versions, row.names = FALSE)

cat("\nTimings:\n")
knitr::kable(tab, row.names = FALSE, digits = 3)

cat("\nSingle-thread overhead check (per workload):\n")
ov <- merge(baseline_sys, baseline_rdevel, by = "workload", all = TRUE)
ov <- merge(ov, baseline_mtl, by = "workload", all = TRUE)
ov$ratio_mtl_vs_sys <- ov$baseline_mtl_lapply / ov$baseline_sys_lapply
ov$ratio_mtl_vs_rdevel <- ov$baseline_mtl_lapply / ov$baseline_rdevel_lapply
knitr::kable(
  ov[, c("workload", "ratio_mtl_vs_sys", "ratio_mtl_vs_rdevel")],
  row.names = FALSE,
  digits = 3
)

cat("\nScaling (speedup vs `mtl` lapply):\n")
mtl_only <- tab_plot[tab_plot$build == "mtl", , drop = FALSE]
mtl_only$threads_plot <- ifelse(mtl_only$label == "lapply", 1L, mtl_only$threads)
mtl_only$threads_plot <- as.integer(mtl_only$threads_plot)
mtl_only <- mtl_only[order(mtl_only$workload, mtl_only$threads_plot), ]

cols <- c("etl_group_mean" = "#1b9e77", "cos_seq" = "#d95f02", "alloc_pressure" = "#7570b3")
u <- unique(mtl_only$workload)

plot(
  NA, xlim = range(mtl_only$threads_plot), ylim = c(0, max(mtl_only$speedup_vs_mtl_lapply, na.rm = TRUE)),
  xlab = "threads", ylab = "speedup vs mtl lapply", xaxt = "n"
)
axis(1, at = sort(unique(mtl_only$threads_plot)))
for (nm in u) {
  d <- mtl_only[mtl_only$workload == nm, ]
  lines(d$threads_plot, d$speedup_vs_mtl_lapply, type = "b", pch = 19, col = cols[[nm]])
}
legend("topleft", legend = u, col = cols[u], lty = 1, pch = 19, bty = "n")

cat("\nbench::mark (mtl build, same workload; plotted):\n")
suppressPackageStartupMessages(library(bench))
print(bm$bench)
plot(bm$bench)
```

## Notes / Limitations

- Workloads that hammer global mutable runtime structures (notably string/symbol interning) may scale poorly.
- Some operations are still serialized under a global lock when run from workers (notably selected native interfaces), to avoid corrupting global process state.
- This is runtime work: correctness and safety come before "make everything parallel".
