---
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 4
)
```

# Threaded R Runtime (Experimental)

This branch experiments with evaluating R code on multiple OS threads inside one process.

Current user-facing APIs:

- `mtlapply(X, FUN, ...)`: threaded apply with the same signature as `lapply()`.
- `background(expr, env = parent.frame())`: enqueue work on the shared thread pool.
- `wait(futures, timeout = Inf)`: pop the next completed future.
- `cancel(future)`: request cancellation for pending/running futures.

Thread-pool size is controlled globally with:

```r
options(mtlapply.threads = 8L)
```

## Quick Start

```r
options(mtlapply.threads = 8L)

# 1) mtlapply
A <- matrix(runif(1000 * 1000), 1000, 1000)
B <- matrix(runif(1000 * 1000), 1000, 1000)
system.time(lapply(1:20, \(i) (A %*% B) + i))
system.time(mtlapply(1:20, \(i) (A %*% B) + i))

# 2) background + wait
work <- function(i) {
  x <- as.double(i)
  for (k in seq_len(500000L)) {
    x <- x + sin(k + x) + cos(k - x * 0.5)
  }
  x
}

pending <- lapply(1:32, \(i) background(work(i)))
out <- vector("list", length(pending))
ids <- seq_along(pending)

while (length(pending)) {
  got <- wait(pending)
  idx <- as.integer(attr(got, "index")[[1L]])
  out[[ids[[idx]]]] <- got$value
  pending <- pending[-idx]
  ids <- ids[-idx]
}
```

## Benchmark Snapshot (Checkpoint 2026-02-12)

The tables below are from concrete benchmark runs in this branch at the
checkpoint tagged on 2026-02-12.

```{r benchmark-snapshot, echo=FALSE}
parity <- data.frame(
  workload = c("etl_group_mean", "cos_seq", "alloc_pressure"),
  rdevel_lapply_s = c(1.095, 0.504, 0.121),
  mtl_lapply_s = c(0.977, 0.426, 0.101),
  ratio_mtl_vs_rdevel = c(0.892, 0.845, 0.835),
  stringsAsFactors = FALSE
)

scale_tab <- data.frame(
  workload = c(
    "alloc_pressure", "alloc_pressure", "alloc_pressure", "alloc_pressure",
    "cos_seq", "cos_seq", "cos_seq", "cos_seq",
    "etl_group_mean", "etl_group_mean", "etl_group_mean", "etl_group_mean"
  ),
  threads = c(1, 2, 4, 8, 1, 2, 4, 8, 1, 2, 4, 8),
  mtlapply_s = c(0.095, 0.114, 0.066, 0.102, 0.410, 0.293, 0.167, 0.167, 0.933, 0.593, 0.304, 0.153),
  lapply_s = c(0.101, 0.101, 0.101, 0.101, 0.426, 0.426, 0.426, 0.426, 0.977, 0.977, 0.977, 0.977),
  speedup = c(1.063, 0.886, 1.530, 0.990, 1.039, 1.454, 2.551, 2.551, 1.047, 1.648, 3.214, 6.386),
  efficiency = c(1.063, 0.443, 0.383, 0.124, 1.039, 0.727, 0.638, 0.319, 1.047, 0.824, 0.803, 0.798),
  stringsAsFactors = FALSE
)

perf_matmul <- data.frame(
  case = c("matmul_1000_n20", "matmul_100_n20000"),
  rows = c(1000, 100),
  cols = c(1000, 100),
  n = c(20, 20000),
  threads = c(8, 8),
  reps = c(3, 3),
  lapply_median_s = c(5.249, 6.386),
  mtlapply_median_s = c(0.805, 0.811),
  speedup = c(6.520, 7.874),
  efficiency = c(0.815, 0.984),
  stringsAsFactors = FALSE
)

bg_burst <- data.frame(
  mode = c("background", "background", "background", "serial"),
  threads = c(2, 4, 8, 1),
  elapsed_s = c(0.280, 0.280, 0.279, 0.620),
  throughput_req_s = c(342.857, 342.857, 344.086, 154.839),
  p50_s = c(0.152, 0.152, 0.152, 0.313),
  p95_s = c(0.277, 0.277, 0.278, 0.590),
  p99_s = c(0.280, 0.280, 0.278, 0.613),
  speedup_vs_serial = c(2.214, 2.214, 2.222, 1.000),
  p95_gain_vs_serial = c(2.126, 2.126, 2.121, 1.000),
  stringsAsFactors = FALSE
)

shiny_full <- data.frame(
  mode = c("sync", "threadpool"),
  sessions_total = c(8, 8),
  sessions_completed = c(8, 8),
  sessions_failed = c(0, 0),
  median_total_s = c(175.144, 23.140),
  median_busy_s = c(153.024, 1.024),
  p95_busy_s = c(153.366, 1.137),
  throughput_completed_sess_per_s = c(0.046, 0.332),
  stringsAsFactors = FALSE
)
```

## 1) Serial Parity: `lapply` vs `R-devel`

Goal: no single-thread slowdown for normal serial code.

```{r serial-parity, echo=FALSE}
knitr::kable(parity, digits = 3, row.names = FALSE)
```

Interpretation:

- `ratio_mtl_vs_rdevel <= 1` means parity or better.
- In this checkpoint artifact, all listed workloads are at or faster than `R-devel` for `lapply`.

## 2) `mtlapply` Scaling and Efficiency

Speedup baseline here is `mtl` build `lapply` on the same workload.
Efficiency is `speedup / threads`.

```{r mtlapply-scaling, echo=FALSE}
knitr::kable(scale_tab, digits = 3, row.names = FALSE)
```

```{r mtlapply-speedup-plot, echo=FALSE}
cols <- c("etl_group_mean" = "#1b9e77", "cos_seq" = "#d95f02", "alloc_pressure" = "#7570b3")
plot(
  NA,
  xlim = range(scale_tab$threads),
  ylim = c(0, max(scale_tab$speedup)),
  xlab = "threads",
  ylab = "speedup vs lapply (same build)",
  xaxt = "n"
)
axis(1, at = sort(unique(scale_tab$threads)))
for (nm in unique(scale_tab$workload)) {
  d <- scale_tab[scale_tab$workload == nm, ]
  lines(d$threads, d$speedup, type = "b", pch = 19, col = cols[[nm]])
}
abline(a = 0, b = 1, lty = 3, col = "gray40")
legend("topleft", legend = unique(scale_tab$workload), col = cols[unique(scale_tab$workload)], lty = 1, pch = 19, bty = "n")
```

Workload behavior from this run:

- `etl_group_mean`: near-linear, strong scaling to 8 threads.
- `cos_seq`: scales through 4 threads, then plateaus at 8 threads.
- `alloc_pressure`: weak scaling; expected for allocation/GC-heavy closures.

## 3) Dense Numeric Workload (Matrix Multiply Loop)

This comes from `tools/mtl-threadpool-perf-smoke.R` with 8 threads.

```{r matmul-perf, echo=FALSE}
knitr::kable(perf_matmul, digits = 3, row.names = FALSE)
```

At this checkpoint:

- `matmul_1000_n20`: 6.52x speedup, 81.5% efficiency.
- `matmul_100_n20000`: 7.87x speedup, 98.4% efficiency.

## 4) `background()` / `wait()` Burst Throughput

This is a Shiny-style request burst simulation from
`tools/mtl-shiny-background-smoke.R`.

```{r background-wait-perf, echo=FALSE}
knitr::kable(bg_burst, digits = 3, row.names = FALSE)
```

For the 8-thread row in this checkpoint:

- Throughput speedup vs serial: ~2.22x.
- P95 gain vs serial: ~2.12x.
- Efficiency (speedup / 8): ~27.8%.

This reflects queueing/coordination overhead and main-thread orchestration in
the current `background()/wait()` path.

## 5) End-to-End Shiny Load Test (Full Recording, No Failures)

From `tools/shiny-threadpool-bench`:

- sync app: `label_sync`
- threadpool app: `label_mt` (`ExtendedTask(offload = "threadpool")`)
- `workers = 8`
- `SHINY_BENCH_WORK_SCALE = 300`
- full recording replay (`recording.log`)

```{r shiny-full, echo=FALSE}
knitr::kable(shiny_full, digits = 3, row.names = FALSE)

sync <- shiny_full[shiny_full$mode == "sync", ]
mtp <- shiny_full[shiny_full$mode == "threadpool", ]
summary_tab <- data.frame(
  metric = c("median_total_s", "median_busy_s", "p95_busy_s", "throughput_completed_sess_per_s"),
  sync = c(sync$median_total_s, sync$median_busy_s, sync$p95_busy_s, sync$throughput_completed_sess_per_s),
  threadpool = c(mtp$median_total_s, mtp$median_busy_s, mtp$p95_busy_s, mtp$throughput_completed_sess_per_s),
  gain = c(
    sync$median_total_s / mtp$median_total_s,
    sync$median_busy_s / mtp$median_busy_s,
    sync$p95_busy_s / mtp$p95_busy_s,
    mtp$throughput_completed_sess_per_s / sync$throughput_completed_sess_per_s
  )
)
knitr::kable(summary_tab, digits = 3, row.names = FALSE)
```

For this run, both modes completed with zero failures (`sessions_failed = 0`),
and the threadpool app improved both latency and throughput materially.

## Reproducing These Benchmarks

```sh
# Serial parity + mtlapply scaling artifacts
R --vanilla -q -f bench/readme_bench_run.R --args bench/results/system_latest.rds
/usr/local/bin/R-devel --vanilla -q -f bench/readme_bench_run.R --args bench/results/rdevel_latest.rds
build-mtl-shlib/bin/R --vanilla -q -f bench/readme_bench_run.R --args bench/results/mtl_latest.rds

# Matrix benchmark
build-mtl-shlib/bin/R --vanilla -q -f tools/mtl-threadpool-perf-smoke.R --args 8 3 0.45

# background()/wait() burst benchmark
build-mtl-shlib/bin/R --vanilla -q -f tools/mtl-shiny-background-smoke.R --args \
  bench/results/mtl_shiny_background_smoke_checkpoint.csv \
  bench/figures/mtl_shiny_background_smoke_checkpoint.png \
  96 3 2,4,8 1.5 120000
```

## Guardrails and Unsupported Patterns (Current)

`mtlapply()` workers are read-mostly with respect to process-global state.

- Writes to `globalenv()` from worker threads are rejected.
- Worker writes to `options(mtlapply.threads = ...)` or `options(threads = ...)` are rejected.
- Worker `options()` changes are job-local and do not propagate to the main thread.
- Some operations with shared process-wide state may serialize under an internal global lock.
- Worker registration of R-level finalizer functions is not supported.
- JIT compilation of closures is currently disabled in worker threads.

In practice: pure/mostly-local closure workloads scale well; global-state and
allocation-heavy workloads can flatten or regress.
