---
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 4
)
```

# Threaded R Runtime (Experimental)

This branch experiments with evaluating R code on multiple OS threads inside one process.

Current user-facing APIs:

- `mtlapply(X, FUN, ...)`: threaded apply with the same signature as `lapply()`.
- `background(expr, env = parent.frame())`: enqueue work on the shared thread pool.
- `wait(futures, timeout = Inf)`: pop the next completed future.
- `cancel(future)`: request cancellation for pending/running futures.

Thread-pool size is controlled globally with:

```r
options(mtlapply.threads = 8L)
```

## Quick Start

```r
options(mtlapply.threads = 8L)

# 1) mtlapply
A <- matrix(runif(1000 * 1000), 1000, 1000)
B <- matrix(runif(1000 * 1000), 1000, 1000)
system.time(lapply(1:20, \(i) (A %*% B) + i))
system.time(mtlapply(1:20, \(i) (A %*% B) + i))

# 2) background + wait
work <- function(i) {
  x <- as.double(i)
  for (k in seq_len(500000L)) {
    x <- x + sin(k + x) + cos(k - x * 0.5)
  }
  x
}

pending <- lapply(1:32, \(i) background(work(i)))
out <- vector("list", length(pending))
ids <- seq_along(pending)

while (length(pending)) {
  got <- wait(pending)
  idx <- as.integer(attr(got, "index")[[1L]])
  out[[ids[[idx]]]] <- got$value
  pending <- pending[-idx]
  ids <- ids[-idx]
}
```

## Benchmark Snapshot (Committed Artifacts)

The tables below are generated from committed artifacts in `bench/results/`.
Regenerate them with `tools/mtl-bench-refresh.sh`, then inspect changes with
`git diff`.

```{r benchmark-snapshot, echo=FALSE}
read_required <- function(path) {
  if (!file.exists(path)) {
    stop(
      sprintf(
        "missing benchmark artifact: %s\nrun tools/mtl-bench-refresh.sh first",
        path
      ),
      call. = FALSE
    )
  }
}

read_results <- function(path) {
  read_required(path)
  x <- readRDS(path)
  stopifnot(is.list(x), is.data.frame(x$results))
  x$results
}

system_res <- read_results("bench/results/system_latest.rds")
rdevel_res <- read_results("bench/results/rdevel_latest.rds")
mtl_res <- read_results("bench/results/mtl_latest.rds")

lapply_join <- function(lhs, rhs, lhs_name, rhs_name) {
  l <- subset(lhs, method == "lapply", c("workload", "median_seconds"))
  r <- subset(rhs, method == "lapply", c("workload", "median_seconds"))
  names(l)[2] <- lhs_name
  names(r)[2] <- rhs_name
  out <- merge(l, r, by = "workload", all = FALSE, sort = TRUE)
  out
}

parity <- lapply_join(mtl_res, rdevel_res, "mtl_lapply_s", "rdevel_lapply_s")
parity$ratio_mtl_vs_rdevel <- parity$mtl_lapply_s / parity$rdevel_lapply_s
parity <- parity[, c("workload", "rdevel_lapply_s", "mtl_lapply_s", "ratio_mtl_vs_rdevel")]

ml <- subset(mtl_res, method == "lapply", c("workload", "median_seconds"))
names(ml)[2] <- "lapply_s"
mm <- subset(mtl_res, method == "mtlapply", c("workload", "threads", "median_seconds"))
names(mm)[3] <- "mtlapply_s"
scale_tab <- merge(mm, ml, by = "workload", all.x = TRUE, sort = TRUE)
scale_tab$speedup <- scale_tab$lapply_s / scale_tab$mtlapply_s
scale_tab$efficiency <- scale_tab$speedup / scale_tab$threads
scale_tab <- scale_tab[, c("workload", "threads", "mtlapply_s", "lapply_s", "speedup", "efficiency")]
scale_tab <- scale_tab[order(scale_tab$workload, scale_tab$threads), ]

read_required("bench/results/threadpool_perf_checkpoint.csv")
perf_matmul <- read.csv("bench/results/threadpool_perf_checkpoint.csv", stringsAsFactors = FALSE)

read_required("bench/results/mtl_shiny_background_smoke_checkpoint.csv")
bg_burst <- read.csv("bench/results/mtl_shiny_background_smoke_checkpoint.csv", stringsAsFactors = FALSE)

shiny_full <- if (file.exists("bench/results/shiny_scale300_w8_full.csv")) {
  read.csv("bench/results/shiny_scale300_w8_full.csv", stringsAsFactors = FALSE)
} else {
  data.frame()
}
```

## 1) Serial Parity: `lapply` vs `R-devel`

Goal: no single-thread slowdown for normal serial code.

```{r serial-parity, echo=FALSE}
knitr::kable(parity, digits = 3, row.names = FALSE)
```

Interpretation:

- `ratio_mtl_vs_rdevel <= 1` means parity or better.
- In this checkpoint artifact, all listed workloads are at or faster than `R-devel` for `lapply`.

## 2) `mtlapply` Scaling and Efficiency

Speedup baseline here is `mtl` build `lapply` on the same workload.
Efficiency is `speedup / threads`.

```{r mtlapply-scaling, echo=FALSE}
knitr::kable(scale_tab, digits = 3, row.names = FALSE)
```

```{r mtlapply-speedup-plot, echo=FALSE}
cols <- c("etl_group_mean" = "#1b9e77", "cos_seq" = "#d95f02", "alloc_pressure" = "#7570b3")
plot(
  NA,
  xlim = range(scale_tab$threads),
  ylim = c(0, max(scale_tab$speedup)),
  xlab = "threads",
  ylab = "speedup vs lapply (same build)",
  xaxt = "n"
)
axis(1, at = sort(unique(scale_tab$threads)))
for (nm in unique(scale_tab$workload)) {
  d <- scale_tab[scale_tab$workload == nm, ]
  lines(d$threads, d$speedup, type = "b", pch = 19, col = cols[[nm]])
}
abline(a = 0, b = 1, lty = 3, col = "gray40")
legend("topleft", legend = unique(scale_tab$workload), col = cols[unique(scale_tab$workload)], lty = 1, pch = 19, bty = "n")
```

Workload behavior from this run:

- `etl_group_mean`: near-linear, strong scaling to 8 threads.
- `cos_seq`: scales through 4 threads, then plateaus at 8 threads.
- `alloc_pressure`: weak scaling; expected for allocation/GC-heavy closures.

## 3) Dense Numeric Workload (Matrix Multiply Loop)

This comes from `tools/mtl-threadpool-perf-smoke.R` with 8 threads.

```{r matmul-perf, echo=FALSE}
knitr::kable(perf_matmul, digits = 3, row.names = FALSE)
```

At this checkpoint:

- `matmul_1000_n20`: 6.52x speedup, 81.5% efficiency.
- `matmul_100_n20000`: 7.87x speedup, 98.4% efficiency.

## 4) `background()` / `wait()` Burst Throughput

This is a Shiny-style request burst simulation from
`tools/mtl-shiny-background-smoke.R`.

```{r background-wait-perf, echo=FALSE}
knitr::kable(bg_burst, digits = 3, row.names = FALSE)
```

For the 8-thread row in this checkpoint:

- Throughput speedup vs serial: ~2.22x.
- P95 gain vs serial: ~2.12x.
- Efficiency (speedup / 8): ~27.8%.

This reflects queueing/coordination overhead and main-thread orchestration in
the current `background()/wait()` path.

## 5) End-to-End Shiny Load Test (Full Recording, No Failures)

From `tools/shiny-threadpool-bench`:

- sync app: `label_sync`
- threadpool app: `label_mt` (`ExtendedTask(offload = "threadpool")`)
- `workers = 8`
- `SHINY_BENCH_WORK_SCALE = 300`
- full recording replay (`recording.log`)

```{r shiny-full, echo=FALSE}
if (!nrow(shiny_full)) {
  cat("`bench/results/shiny_scale300_w8_full.csv` not found.\n")
  cat("Run the shiny loadtest workflow and commit the artifact to populate this section.\n")
} else {
  knitr::kable(shiny_full, digits = 3, row.names = FALSE)

  sync <- shiny_full[shiny_full$mode == "sync", ]
  mtp <- shiny_full[shiny_full$mode == "threadpool", ]
  summary_tab <- data.frame(
    metric = c("median_total_s", "median_busy_s", "p95_busy_s", "throughput_completed_sess_per_s"),
    sync = c(sync$median_total_s, sync$median_busy_s, sync$p95_busy_s, sync$throughput_completed_sess_per_s),
    threadpool = c(mtp$median_total_s, mtp$median_busy_s, mtp$p95_busy_s, mtp$throughput_completed_sess_per_s),
    gain = c(
      sync$median_total_s / mtp$median_total_s,
      sync$median_busy_s / mtp$median_busy_s,
      sync$p95_busy_s / mtp$p95_busy_s,
      mtp$throughput_completed_sess_per_s / sync$throughput_completed_sess_per_s
    )
  )
  knitr::kable(summary_tab, digits = 3, row.names = FALSE)
}
```

For this run, both modes completed with zero failures (`sessions_failed = 0`),
and the threadpool app improved both latency and throughput materially.

## Reproducing These Benchmarks

```sh
# One-shot refresh of committed benchmark artifacts + markdown outputs
tools/mtl-bench-refresh.sh build-mtl-shlib R /usr/local/bin/R-devel

# Then inspect impact
git diff -- bench/LATEST.md README.md bench/results/
```

## Guardrails and Unsupported Patterns (Current)

`mtlapply()` workers are read-mostly with respect to process-global state.

- Writes to `globalenv()` from worker threads are rejected.
- Worker writes to `options(mtlapply.threads = ...)` or `options(threads = ...)` are rejected.
- Worker `options()` changes are job-local and do not propagate to the main thread.
- Some operations with shared process-wide state may serialize under an internal global lock.
- Worker registration of R-level finalizer functions is not supported.
- JIT compilation of closures is currently disabled in worker threads.

In practice: pure/mostly-local closure workloads scale well; global-state and
allocation-heavy workloads can flatten or regress.
